---
title: "Pre-workshop Installation Instructions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{preworkshop_instructions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Contents borrowed and modified from [UVA's Data Science Essentials in R series](https://uvastatlab.github.io/phdplus/intror.html)

## Before the first session

To participate in the R workshop, please bring a laptop with R and RStudio installed. We recommend that you have the latest version of R (3.6.1), the `tidyverse` package (1.2.1), and the `learnr` package (0.9.2.1). You will then need to use the `devtools` package to download the `workr` package we developed for our workshop tutorials. You need to have RStudio installed, but it is less crucial that you are using the most recent version (1.2.5001).

**Do you already have R and RStudio installed?**  

* No - follow the instructions for “I do not have R installed”  
* Yes - follow the instructions for “I have R installed”

## “I do not have R installed”

You should install R, RStudio, `tidyverse`, and `learnr`.

### Installing R

#### Windows:
1. Go to https://cloud.r-project.org/bin/windows/base/
2. Click the “Download R 3.6.1 for Windows” link.
3. When the file finishes downloading, double-click to install. You should be able to click “Next” to all dialogs to finish the installation.

#### Mac:
1. Go to https://cloud.r-project.org/bin/macosx/
2. Click the link “R-3.6.1.pkg”
3. When the file finishes downloading, double-click to install. You should be able to click “Next” to all dialogs to finish the installation.

#### Linux:
For any adventurous Linux users in our group follow this guide (https://github.com/duckmayr/install-update-r-on-linux) to install/upgrade to the most recent version of R on Ubuntu (18.04) or Mint (19).

### Installing RStudio
1. Go to [the RStudio download page](https://www.rstudio.com/products/rstudio/download/#download).
2. Under “Installers for Supported Platforms” select the appropriate installer for your operating system
3. When the file finishes downloading, double-click to install. You should be able to click “Next” to all dialogs to finish the installation.

### Installing packages
Skip ahead to the **Installing and updating packages** section for instructions on how to install the necessary packages for our workshop.

## “I have R installed”

The workshops run more smoothly when everyone is using the same version of R,  `tidyverse`, and `learnr`. Please update R, `tidyverse`, and `learnr` if necessary (and less crucially, RStudio).

### Verify R version

Open RStudio. At the top of the Console you will see session info. The first line tells you which version of R you are using. If RStudio is already open and you’re deep in a session, type `R.version.string` in the console and enter to print out the R version.

Do you have R version 3.6.1 installed?

* No - follow the instructions for “Updating R”
* Yes - Great! Do you have `tidyverse` and `learnr` installed?
  + No or I don’t know - See “Installing `tidyverse`”
  + Yes - Great! Go to Go to Tools > Check for Package Updates. If there’s an update available for `tidyverse`, install it.

### Updating R/RStudio

#### Windows
To update R on Windows, try using the package `installr` (only for Windows).  

1. Install and load installr:  
```{r eval=FALSE}
install.packages("installr")
library(installr)
```  
2. Call `updateR()` function. This will start the updating process of your R installation by: “finding the latest R version, downloading it, running the installer, deleting the installation file, copy and updating old packages to the new R installation.”  
3. From within RStudio, go to Help > Check for Updates to install newer version of RStudio (if available, optional).

#### Mac
On Mac, you can simply download and install the newest version of R. When you restart RStudio, it will use the updated version of R.  

1. Go to https://cloud.r-project.org/bin/macosx/  
2. Click the link “R-3.6.1.pkg”  
3. When the file finishes downloading, double-click to install. You should be able to click “Next” to all dialogs to finish the installation.  
4. From within RStudio, go to Help > Check for Updates to install newer version of RStudio (if available, optional).  

#### Linux:
Again, for any adventurous Linux users in our group follow this guide (https://github.com/duckmayr/install-update-r-on-linux) to install/upgrade to the most recent version of R on Ubuntu (18.04) or Mint (19).

## Installing and updating packages

### Installing `tidyverse` and `learnr`

1. Open RStudio
2. Go to `Tools` > `Install Packages`
3. Enter `tidyverse`
4. Select `Install`

Follow the same protocol for installing the `learnr` package, but replace `tidyverse` with `learnr`. Alternatively, you can install packages by running the command `install.packages()` in your console. You can install multiple packages as once by combining package names into a vector as follows:


```{r, eval=FALSE}
install.packages(c("tidyverse","learnr"))
```

### Installing `workr` using the `devtools` package

While there are many useful packages hosted on the CRAN (the global repository of R packages queried using `install.packages()`), you may be interested in using packages that are still in development. Many of these packages are shared on GitHub and can be downloaded using `install_github()`, a function of the `devtools` package. Below is an example of how to install the `workr` package using `devtools`.

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("https://github.com/natalieoshea/workr.git") 
## If it asks to install or update packages, be sure to select "CRAN packages only"
```

Including `devtools:::` before `install_github()` allows you to call the function without having to load the entire library. However, if you are planning to use additional functions from the `devtools` package you may want to fully load the package (`library(devtools)`--> `install_github()`)

### Check for package updates

If you already have all packages installed you can check for updates to CRAN packages using the `update.packages()` command or by going to `Tools`--> `Check for Package Updates`. Packages installed from GitHub can be updated using `devtools` with the `update_packages()` command. For example, to check for updates to one of our tutorials, be sure to run `devtools::update_packages(workr)` before you begin.

